Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-03 paper.tex
address some comments and update figures

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from 1000 Genomes and ADRP   \sgcomment{define} illustrate diversity from
        western, eastern and southern Africa. We chose representatives from each region
        to build parameterized models, including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual (bold fonts \sgcomment{I don't see bold fonts}). (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British. \sgcomment{I think we should indicate the origin on PCAS, also possible variance explained.} (D) \texttt{ADMIXTURE}
        illustrates signatures of recent gene flow in Africa which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:1}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fig2}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of concurrence \sgcomment{concurrence or co-occurence? What does this mean in either case?} of contemporaneous populations \sgcomment{Is what we mean: Migration rates are constant over epochs delimited by population splits or mergers, see Supplementary table XX? }. The
        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
        continuous-migration model (A, with $LL=-115,500$ ).
    }
    \label{fig:2}
\end{figure}

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $1e-4$ 
        In both models, the branch ancestral to the Nama shares a common ancestral population 
        with the other African groups $\sim$120--135kya. Following this divergence, 
        the population ancestral to other African groups branches into West and East African
        groups 60kya. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
        and their ancestors, and 2) Nama and Mende, respectively.
        `Divergence times' correspond to the most recent common ancestral population, and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
2 contribution to the Western African Mende resulted in better model fits (LL \sgcomment{Something is missing here}).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (top row:
        continuous migration, bottom row: merger with stem migration), we predicted
       pairwise differences $H_{i,j}$ between individuals sampled from populations $i$ and $j$ 
       existing at time $t$. 
        To understand how drift between stems explains present-day structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of sampled present-day
        populations that aligns with drift between past populations (Here Nama and Mende, see Section~S\ref{sec:f4} for details and additional comparisons \sgcomment{do we have extra comparisons there?}).
        Both models infer deep population structure with modest contributions to present-day genetic differentiation. Most present differentiation dates back to the last 100kya.}
    \label{fig:3}
\end{figure}

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$kya and $100$kya in humans, and thus inferred an increase 
in $N_e$ during the same period. \citep{Li, Durbin}. 
This increase in $N_e$ could be due to either an increase in population or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.

@@

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past ~120kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Nama has an
\sgcomment{XX} probability of being inherited from the ancestral ‘southern’ subpopulation 50kya versus
\sgcomment{XX} probability of being inherited from the ‘eastern’ subpopulation (Figure \sgcomment{XX}). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table 1).

>>>>

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:1}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/fig2-melisa}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of co-existence of contemporaneous populations with constant
        migration rates over those intervals. The
        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
        continuous-migration model (A, with $LL=-115,500$ ).
    }
    \label{fig:2}
\end{figure}

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $1e-4$ 
        In both models, the branch ancestral to the Nama shares a common ancestral population 
        with the other African groups $\sim$120--135kya. Following this divergence, 
        the population ancestral to other African groups branches into West and East African
        groups 60kya. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
        and their ancestors, and 2) Nama and Mende, respectively.
        `Divergence times' correspond to the most recent common ancestral population, and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya.
The later Stem 2 contribution to the Western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains present-day structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled present-day
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        present-day genetic differentiation.
        Most present differentiation dates back to the last 100kya.
    }
    \label{fig:3}
\end{figure}

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$kya and $100$kya in humans, and thus inferred an increase 
in $N_e$ during the same period \citep{Li2011-le}.
This increase in $N_e$ could be due to either an increase in population or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.

@@

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past $\sim120$kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) 150kya versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}).

### 2022-03-03 supplement.tex
address some comments and update figures

The IICR is only a reliable estimator of $N_e$ in the absence of population structure.
However, since IIRC are readily estimated and visually interesting,
it is commonly attempted even when the assumption is not met
\citep{}\aprcomment{what citation did you have in mind here, Simon?}.
Even if the assumption is not met, the inferred IIRC can be interpreted as a
summary statistic for which model predictions can be compared to data.
It can therefore be
informative about aspects of the data that our models fail to predict.
We also note that the coalescence rates inferred from data used a larger set of
populations from the combined 1000 Genomes/AGRP dataset, while our model
simulations only included a subset of 5 populations. This may lead to
differences in resolution of the inferences.

>>>>

The IICR is only a reliable estimator of $N_e$ in the absence of population structure.
However, since IIRC are readily estimated and visually interesting,
it is commonly attempted even when the assumption is not met.
Even if the assumption is not met, the inferred IIRC can be interpreted as a
summary statistic for which model predictions can be compared to data.
It can therefore be
informative about aspects of the data that our models fail to predict.
We also note that the coalescence rates inferred from data used a larger set of
populations from the combined 1000 Genomes/AGRP dataset, while our model
simulations only included a subset of 5 populations. This may lead to
differences in resolution of the inferences.
